---
title: "Probes to genes, using new mapping file"
output: html_notebook
---

You need to run Concat_feature_importances.ipynb before running this.

Parameters to set:
```{r}

cancer_type = ''
min_count = 1 # how many times a feature had to be found in consecutive runs of xgboost to be taken and transformed to a gene
strict_mapping <- FALSE # if true, uses 1500 mapping file, else uses 5000 mapping file. We use the 5000 one.
only_islands <- FALSE # only take probes within CpG islands
near_islands <- FALSE # takes probes within CpG islands or their shore or shelf (so a 4000bp window around either side of each island)

island_label <- ''
if (only_islands == TRUE) {
  island_label <- 'only_islands'
} else if (near_islands == TRUE) {
  island_label <- 'near_islands'
}

# change, depending on the cancer type:
feature_file <- paste('feature_importances/', cancer_type, '_concatenated.csv', sep='') # features found by xgboost
print("Using features from:")
print(feature_file)

if (strict_mapping == TRUE) {
  mapping_file <- '../checking_mapping/my_mapping_1500.rda'
  save_file <- paste('features_as_genes/all_genes_min_count_', min_count,'_', island_label, '_', cancer_type, '_strict.csv', sep = '')
  
} else {
  mapping_file <- '../checking_mapping/my_mapping_5000.rda'  
  save_file <- paste('features_as_genes/all_genes_min_count_', min_count, '_', island_label, '_', cancer_type, '_relaxed.csv', sep = '')
}

```


Get a list of probes, in the order that they are in in the m_value matrices
```{r}

# probes_file <- '../data_preprocessing/dataset/numpy/genes/probes.csv' # a list of the probes, in the order that they are in the m_value matrices
# total_probes <- read.table(probes_file)


# from the original data, read in the probe list:
library(data.table)
if (cancer_type == '') {
  total_probes <- fread('../data_preprocessing/dataset/pandas/m_values/TCGA-all.csv', sep = '\t', select = c(1))  
} else {
  total_probes <- fread(paste('../data_preprocessing/dataset/pandas/m_values/TCGA-', cancer_type, '.csv', sep = ''), sep = '\t', select = c(1), header = TRUE)
}


```


Load new mapping and feature list:
```{r message = FALSE}
load(mapping_file)

features = read.table(feature_file, sep = ',', header = TRUE)

# filter
library(dplyr)
features <- features %>%
  filter(count >= min_count)

# View(features)
```

Features to probes:
```{r}

probes <- total_probes$probe[features$feature + 1] # feature indexing is from 0, but R indexing is from 1, so we add 1 - I have verified this is correct
# View(probes)
```


Functions taken from makeMyMapping.Rmd:
```{r}

# given a probe name, gets the gene ids that are mapped to that probe
get_gene_ids <- function(probe, mapping) {
  rows = mapping[mapping$probe == probe, ]
  ids = rows$ensembl_gene_id
  return(ids)
}

get_CGI_location <- function(probe, mapping) {
  rows = mapping[mapping$probe == probe, ]
  location = rows$feature_type
  if (identical(location, character(0))) return(".") # if probe didn't map to any row, return not at any CGI location
  return(location[[1]]) # just get first location, as the same across all rows
}

get_all_info <- function(probe, mapping) {
  rows = mapping[mapping$probe == probe, ]
  return(rows)
}

```

Mapping probes to genes:
```{r}
library(purrr)

# filter probes based on their CGI loction
if (only_islands == TRUE) {
  location <- map(probes, get_CGI_location, my_mapping)
  probes <- probes[location == "Island"]
} else if (near_islands == TRUE) {
  location <- map(probes, get_CGI_location, my_mapping)
  probes <- probes[location != "."]
}

genes <- map(probes, get_gene_ids, my_mapping)

# genes

genes <- unlist(genes)
# View(genes)
```

Save gene list
```{r}
# genes
write.table(genes, save_file, quote = FALSE, row.names = FALSE, col.names = FALSE)
```


Getting stats about found probes - within CGIs?, within genes?, ...
```{r}
all_info <- map(probes, get_all_info, my_mapping) # NOTE: if a probe maps to more than 1 gene, it will have more than 1 GRanges row


stats <- data.frame(row.names = probes)

stats$before_gene = 0
stats$within_gene = 0
stats$after_gene = 0
stats$N_Shelf = 0
stats$N_Shore = 0
stats$Island = 0
stats$S_Shore = 0
stats$S_Shelf = 0
stats$genes_mapped = 0 


# stats['cg06804091', 'within_gene'] <- 0

for (l in all_info) { # for each found probe and its info
  if (length(l) > 0) {
    stats[l[1, ]$probe, "genes_mapped"] <- length(l)
    for (i in 1:length(l)) {
      # print(i)
      # print(l[i, ])
      
      # Do CGIs
      if (l[i, ]$feature_type != '.') {
        val <- stats[l[i, ]$probe, l[i, ]$feature_type]
        stats[l[i, ]$probe, l[i, ]$feature_type] <- val + 1
      }
      
      # Do genes
      if (countOverlaps(l[i, ]$gene_feature.ranges, ranges(l[i, ])) > 0) { # within gene
        stats[l[i, ]$probe, "within_gene"] <- stats[l[i, ]$probe, "within_gene"] + 1
      }
      else {
        if (start(l[i, ]$gene_feature.ranges) > start(l[i, ])) {
          if (as.vector(l[i, ]$gene_feature.strand)[[1]] == "+") {
            stats[l[i, ]$probe, "before_gene"] <- stats[l[i, ]$probe, "before_gene"] + 1
          }
          else { # different for negative strand
            stats[l[i, ]$probe, "after_gene"] <- stats[l[i, ]$probe, "after_gene"] + 1
          }
        }
        
        if (end(l[i, ]$gene_feature.ranges) < start(l[i, ])) {
          if (as.vector(l[i, ]$gene_feature.strand)[[1]] == "+") {
            stats[l[i, ]$probe, "after_gene"] <- stats[l[i, ]$probe, "after_gene"] + 1
          }
          else { # different for negative strand
            stats[l[i, ]$probe, "before_gene"] <- stats[l[i, ]$probe, "before_gene"] + 1
          }
        }
      }
    }
  }
}



# check: (should be 1)
mean(stats$within_gene + stats$before_gene + stats$after_gene == stats$genes_mapped)
```



Visualising stats:
```{r}
library(tidyverse)

# as_tibble(stats) %>%
#   ggplot() +
#   geom_point(mapping = aes(x = 0, y = ifelse(stats$before_gene != 0, stats$before_gene, NA)), position = position_jitter(w = 0.1, h = 0)) + 
#   geom_point(mapping = aes(x = 1, y = ifelse(stats$within_gene != 0, stats$within_gene, NA)), position = position_jitter(w = 0.1, h = 0)) + 
#   geom_point(mapping = aes(x = 2, y = ifelse(stats$after_gene != 0, stats$after_gene, NA)), position = position_jitter(w = 0.1, h = 0))

# total numbers of probes mapped to specific areas. Look at genes and CGI positions separately (on one graph for ease)
total_nums <- colSums(stats)
nums <- as_tibble(total_nums)
nums$names <- names(total_nums)
nums$names <- factor(nums$names, levels = nums$names)
nums %>%
  ggplot() + 
  geom_col(mapping = aes(x = names, y =value))


# how many genes does each probe map to?
as_tibble(stats) %>%
  ggplot() +
  # geom_point(mapping = aes(x = 0, y = genes_mapped), position = position_jitter(w = 0.1, h = 0))
  geom_histogram(mapping = aes(x = genes_mapped), bins = 100)

```

Saving stats:
```{r}
stats['totals', ] = colSums(stats)
stats

if (strict_mapping == TRUE) {
  savepath <- paste('feature_stats/stats_min_count_', min_count, '_', island_label, '_', cancer_type, '_strict.csv', sep = '')
} else {
  savepath <- paste('feature_stats/stats_min_count_', min_count, '_', island_label, '_', cancer_type, '_relaxed.csv', sep = '')
}


write.table(stats, savepath, quote = FALSE, row.names = TRUE, col.names = TRUE)

```

